---
title: "A gentle Introduction to S3 classes"
author: "Edwin Thoen"
date: "8 Dec 2017"
output: html_document
---

Every R user has used S3 classes, most likely even the very first time. Have you ever used a `data.frame`, `factor` or a `Date` variable? Those are just a few examples of them. Still, many don't have a grasp of exactly they work. Functional programming comes more natural to most R users than object oriented programming. However, having some understanding of S3 classes will enlarge  understanding of the objects you use and will expand R programming skills. This is not a thorough introduction into object oriented programming, but rather explores the basics of S3 classes. 

## Types and Classes

R has just a few basic building blocks on which many more complex constructs are built. The most important for now are the vectors `character`, `logical`, `integer`, `double` and `list`. 

```{r}
typeof(1:3)
typeof(list(a = 1:3, b = c("a", "b")))
```

So the function `typeof` returns the underlying type of an object. An S3 object is created by adding a class to a raw object (or to another S3 object). This class is than an attribute of the object.

```{r}
f <- factor(c("a", "b", "c"))
attributes(f)
```

So a `factor` has two attributes, its levels and the class factor. It is built on top of an integer.

```{r}
typeof(f)
```

