---
layout: post
categories: blog
title: "Make your own color palettes with paletti"
base-url: https://EdwinTh.github.io
date: "2017-12-22 16:00:00"
output: html_document
tags: [R, ggplot2, color palette, paletti]
---

Last week I blogged about the [dutchmasters color palettes package](https://edwinth.github.io/blog/dutch-masters/), which was inspired 
by the wonderful [ochRe package](https://github.com/ropenscilabs/ochRe). As 
mentioned I shamelessly copied the package. I replaced the list with character 
vectors containing hex colors and did a find and replace to make it 
`dutchmasters` instead of `ochRe`. This was pretty ugly.  I realized that when 
we would refactor the  `ochRe` functions, thus creating functions that create the functions, there would no longer be a need to copy-paste and find-and-replace. So that is what I did. I refactored and expanded `ochRe`'s chore into `paletti`. (Name chosen because I liked the ring of it).

## Single palettes

`paletti` takes both single palettes (character vectors with hex codes) and 
lists with palettes, like the `ochRe` and `dutchmasters` lists. Lets start with 
a single palette, this might be useful when you want your coporate identity 
colors translated into R. Here I just pick some colors found on the [interweb](http://blog.visme.co/website-color-schemes/)

```{r}
library(paletti)
library(ggplot2)
mycols <- c(
  red    = "#E27D60",
  blue   = "#085DCB",
  orange = "#E8A87C",
  purple = "#C38D9E",
  green  = "#41B3A3"
)
viz_palette(mycols)
```

Now, ochRe provided us with two functions, one two create a `ggplot` scale for 
colours and to create one for fills. These functions can be created in the 
following fashion

```{r}
mycols_fill  <- get_scale_fill(get_pal(mycols))
mycols_color <- get_scale_color(get_pal(mycols))
```

Both now can be used in `ggplot`

```{r}
mtcars$cyl <- as.character(mtcars$cyl)
col_plot <- ggplot(mtcars, aes(mpg, drat, color = cyl)) +
  geom_point(size = 4) 
col_plot + mycols_color()

fill_plot <- ggplot(mtcars, aes(cyl, fill = cyl)) + 
  geom_bar()
fill_plot + mycols_fill()
```

Now, I said I expanded the `ochRe` code a bit. The function `get_hex` will 
produce a function that will return a function in which you can directly return
the hex code by typing its unquoted name. Handy if you want an exact color from
your palette. Prerequisit is that your palette is a named character vector.

```{r}
mycols_hex <- get_hex(mycols)
fill_plot + scale_fill_manual(values = mycols_hex(blue, purple, red))
```

## Multiple palettes in a list

Both `ochRe` and `dutchmasters` offer multiple palettes in a list. The only 
difference from a single palette is that in the returned function you have to 
specify the name of the palette youw want to use. If you don't, it defaults to 
the first palette in the list.

```{r}
ochRe_col         <- get_scale_color(get_pal(ochre_palettes))
dutchmasters_fill <- get_scale_fill(get_pal(dutchmasters))

col_plot + ochRe_col("lorikeet")
fill_plot + dutchmasters_fill("view_of_Delft")
```

And the same holds for the `get_hex` function. You can feed a list with palettes
as well. Note that the palettes that you are going to call must have named
elements.

```{r}
dutchmasters_hex <- get_hex(dutchmasters)
fill_plot + scale_fill_manual(values = dutchmasters_hex("anatomy",
                              black(cloak), red(muscle), brown(table)))
```

That's it, off you go! Add your own color palette(s) and start plotting. Once 
again a major thanks to the ochRe team for the inspiration and the foundations 
on which `paletti` is built.
